---
title: "Homework_Bayesian_networks"
author: "Nikhileswar mada"
date: '`r Sys.Date()`'
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Question 1

### (a)

```{r}

library(gRain)
library(Rgraphviz)
library(gRbase)
library(ggm)
library(graph)
library(deal)
library(bnlearn)
data("ksl")
data<-ksl[,1:8]
head(data)
```
```{r}
unique(data[,c('Hyp')])
unique(data[,c('Smok')])
unique(data[,c('Alc')])
unique(data[,c('Work')])
unique(data[,c('Sex')])
```


Discretizing the log body Mass Index into two classes because there are only two factors or two integers for all the remaining continuous values


```{r}
vars<-c("FEV","Kol","logBMI")
data$FEV<-as.numeric(data$FEV)
data$Kol<-as.numeric(data$Kol)
data[,vars]<-discretize(data[,vars],method='interval',breaks=2)
```

Converting the data from int to numeric to create two numeric variables FEV and Kol along with logBMI which was already in numeric.
Then we applied discretize function in interval of 2 since most other variables have 2 states.

### (b)

```{r}
data <- as.data.frame(lapply(data, as.factor))
head(data)
```


Converting all the variables to factors then we are applying hc from bnlearn library function to generate the bayesian network.


```{r}
library(bnlearn)
model <- hc(data, score = "bic")
net<-as(amat(model),"graphNEL")
plot(net)

```

The structure shows that Hyp is the primary parent of all then it's child are logBMI and Smok and no other variable is depended on the logBMI.This is the dag

```{r}
print(model)
```



The model summary is listed above indicating the relationship between each node and the neighbourhood size along with the branching factor.

### (c)

```{r}
cpdag<-cpdag(model)
```



```{r}
graphviz.plot(cpdag)
```

This is the cpdag of the bayesian network.

### (d)

```{r}
dag <- cextend(cpdag, strict = TRUE)
graphviz.plot(dag)
```
All the v-structures were maintained this shows that the dag here is an equivalence of the dag in b.


### (e)

```{r}
fit<-bn.fit(model,data=data)
fit$FEV
```

This shows that the smoking alone increases the chance of higher forced ejection volumes.Based on the conditional probability tables shows that 1 level of smoking has 75% lower FEV interval whereas 2 level of smoking shows a 67% lower FEV interval meaning a 8% percent increase in higher level of FEV.

Alcohol doesn't affect the Forced ejection volumes since it has d-seperation between them.


### (f)

Since 

```{r}
junction <- compile(as.grain(fit))
jsex <- setEvidence(junction, nodes = c("Alc","Hyp"), states = c('2','1'))
print(querygrain(jsex, nodes = "FEV")$FEV)
```


This shows a lower FEV.


### (g)

Let's use a different discretize model hartemink of two instead of interval of two breaks and see the results.

```{r}
data2<-data(ksl)
data2<-ksl[,1:8]
vars<-c("FEV","Kol","logBMI")
data2$FEV<-as.numeric(data2$FEV)
data2$Kol<-as.numeric(data2$Kol)
data2[,vars]<-discretize(data2[,vars],method='hartemink',breaks=2)
head(data2)
```



Let's look at the dag and cpdag 

```{r}
data2 <- as.data.frame(lapply(data2, as.factor))
model2 <- hc(data2, score = "bic")
net2<-as(amat(model2),"graphNEL")
plot(net2)
```



This shows that the FEV is parent of all and doesn't depend upon any factors unlike the previous dag created using interval.Whereas Hyp is a leaf node here whereas earlier it is base parent.


Let's look at the cpdag

```{r}

cpdag2<-cpdag(model2)
graphviz.plot(cpdag2)
```
The cpdag of two are quite different and the equivalence classes of one is not going to be the other.

\newpage

# Question 2


### (a)

```{r}

marks.dag = empty.graph(nodes = c("MECH", "LAT", "ANL", "VECT", "ALG", "STAT"))
marks.dag = set.arc(marks.dag, from = "MECH", to = "LAT") 
marks.dag = set.arc(marks.dag, from = "ANL", to = "LAT") 
marks.dag = set.arc(marks.dag, from = "LAT", to = "ALG") 
marks.dag = set.arc(marks.dag, from = "LAT", to = "STAT") 
marks.dag = set.arc(marks.dag, from = "LAT", to = "VECT")

graphviz.plot(marks.dag)
```

The dag is created as per the question.

### (b)

Creating the cpdag of the above question.

```{r}
graphviz.plot(cpdag(marks.dag))
```



The cpdag looks the same since changing the directions would cause the dag to create more v-structures.

### (c)

```{r}

hc <- hc(marks)
graphviz.plot(hc)

```

The most likely structure need not be a DAG meaning a directed acyclic graph where our hc structure has cyclic structures.

LAT variable was removed from the structure.

\newpage

# Question 3

## a)

```{r}
library(gRbase)
library(rbmn)
data(carcass)
head(carcass)

carcass$Fat11<-as.numeric(carcass$Fat11)
carcass$Meat11<-as.numeric(carcass$Meat11)
carcass$Fat12<-as.numeric(carcass$Fat12)
carcass$Meat12<-as.numeric(carcass$Meat12)
carcass$Fat13<-as.numeric(carcass$Fat13)
carcass$Meat13<-as.numeric(carcass$Meat13)

# Set the variable Lean Meat to be at the bottom of the network.
nodes <- names(carcass)
nodes <- nodes[nodes != "LeanMeat"]
nodes <- append(nodes, "LeanMeat")

#carcass <- as.data.frame(lapply(carcass, as.factor))

head(carcass)


bl<-matrix(c("LeanMeat", "Fat12", "LeanMeat", "Fat13"), ncol = 2, byrow = TRUE)
bn.hc <- hc(carcass[, nodes],score='bic-g',blacklist=bl)

graphviz.plot(bn.hc)
```

Here the leanMeat is the leaf node and at the bottom of the network.

Taking the model and converting to bayesian network using modelstring developed from

```{r}
bn.hc
```

```{r}
ans<-model2network("[Fat11][Meat11][Fat12|Fat11][Meat12|Meat11][Fat13|Fat11:Meat11:Fat12][Meat13|Fat11:Meat11:Fat12:Meat12][LeanMeat|Fat11:Meat11:Fat12:Fat13]")

bn.learn<-bn.fit(ans,data=carcass)
graphviz.plot(bn.learn)
```

### (b)

Based on the conditional independence test using iamb

```{r}
ans<-rbn(bn.learn,n=10000)
str<-iamb(ans,test="cor",blacklist=bl)
graphviz.plot(str)
```


## c)

The networks are only similar in terms of bottom of the network which is LeanMeat and Fat11 being at the top rest are jumbled across the networks.
The number of v-structures differed.

## d)

```{r}
ans4<-rbn(bn.learn,n=25)
str4<-iamb(ans4,test="cor")
graphviz.plot(str4)
```


The network could not form completely due to low sample size and most are undirected edges.This clearly very different from a and b.

## e)

```{r}
ans5<-rbn(bn.learn,n=100)
str5<-iamb(ans5,test="cor")
graphviz.plot(str5)
```

This clearly got better than the 25 samples but the directions hasn't formed yet and few more samples this would be complete.

\newpage

# Question 4

```{r}
dag <- empty.graph(nodes = c('BirthAsphyxia', 'Disease', 'Age', 'LVH',
'DuctFlow', 'CardiacMixing', 'LungParench', 'LungFlow', 'Sick', 'HypDistrib', 'HypoxiaInO2', 'CO2','ChestXray', 'Grunting', 'LVHreport', 'LowerBodyO2', 'RUQO2', 'CO2Report', 'XrayReport', 'GruntingReport'))
dag <- set.arc(dag, from = 'BirthAsphyxia', to = 'Disease') 
dag <- set.arc(dag, from = 'Disease', to = 'Age')
dag <- set.arc(dag, from = 'Disease', to = 'LVH')
dag <- set.arc(dag, from = 'Disease', to = 'DuctFlow')
dag <- set.arc(dag, from = 'Disease', to = 'CardiacMixing') 
dag <- set.arc(dag, from = 'Disease', to = 'LungParench') 
dag <- set.arc(dag, from = 'Disease', to = 'LungFlow')
dag <- set.arc(dag, from = 'Disease', to = 'Sick')
dag <- set.arc(dag, from = 'LVH', to = 'LVHreport')
dag <- set.arc(dag, from = 'DuctFlow', to = 'HypDistrib')
dag <- set.arc(dag, from = 'CardiacMixing', to = 'HypDistrib') 
dag <- set.arc(dag, from = 'CardiacMixing', to = 'HypoxiaInO2') 
dag <- set.arc(dag, from = 'LungParench', to = 'HypoxiaInO2') 
dag <- set.arc(dag, from = 'LungParench', to = 'CO2')
dag <- set.arc(dag, from = 'LungParench', to = 'ChestXray')
dag <- set.arc(dag, from = 'LungParench', to = 'Grunting')
dag <- set.arc(dag, from = 'LungFlow', to = 'ChestXray')
dag <- set.arc(dag, from = 'Sick', to = 'Grunting')
dag <- set.arc(dag,from = "Sick",to="Age")
dag <- set.arc(dag, from = 'HypDistrib', to = 'LowerBodyO2') 
dag <- set.arc(dag, from = 'HypoxiaInO2', to = 'LowerBodyO2') 
dag <- set.arc(dag, from = 'HypoxiaInO2', to = 'RUQO2')
dag <- set.arc(dag, from = 'CO2', to = 'CO2Report')
dag <- set.arc(dag, from = 'ChestXray', to = 'XrayReport') 
dag <- set.arc(dag, from = 'Grunting', to = 'GruntingReport')
graphviz.plot(dag)

```


```{r}
dag
```



This contains the joint distribution of the dag model inside the model in the form of the modelstring which is 

   [BirthAsphyxia][Disease|BirthAsphyxia][LVH|Disease][DuctFlow|Disease]
   [CardiacMixing|Disease][LungParench|Disease][LungFlow|Disease][Sick|Disease]
   [Age|Disease:Sick][HypDistrib|DuctFlow:CardiacMixing]
   [HypoxiaInO2|CardiacMixing:LungParench][CO2|LungParench]
   [ChestXray|LungParench:LungFlow][Grunting|LungParench:Sick][LVHreport|LVH]
   [LowerBodyO2|HypDistrib:HypoxiaInO2][RUQO2|HypoxiaInO2][CO2Report|CO2]
   [XrayReport|ChestXray][GruntingReport|Grunting]
   
   
### (b)   
   
```{r}

graphviz.plot(moral(dag))
```

### (c)

parameterized using the bn.fit and the method used is maximum likelihood estimate.


```{r}
child<- read.csv("child_network.csv", header=TRUE, colClasses="factor") 
bn.mle <- bn.fit(dag, data = child, method = "mle")
```

### (d)

n13 is the chest X Ray parameter.

The CPT is given as 

```{r}
bn.mle$ChestXray
```


### (e)

n14 is grunting

The CPT of grunting is as follows:

```{r}
bn.mle$Grunting
```


###(f)

Given lowerbodyO2<5 and xray report to be plethoric then based on this we are trying to determine the disease this caused.

```{r}
cpquery(bn.mle, event = (Disease == 'Fallot'), evidence = ((LowerBodyO2 == '<5' & XrayReport == 'Plethoric')))
```

```{r}
cpquery(bn.mle, event = (Disease == 'Lung'), evidence = ((LowerBodyO2 == '<5' & XrayReport == 'Plethoric')))
```

```{r}
cpquery(bn.mle, event = (Disease == 'PAIVS'), evidence = ((LowerBodyO2 == '<5' & XrayReport == 'Plethoric')))
```

```{r}
cpquery(bn.mle, event = (Disease == 'PFC'), evidence = ((LowerBodyO2 == '<5' & XrayReport == 'Plethoric')))
```

```{r}
cpquery(bn.mle, event = (Disease == 'TAPVD'), evidence = ((LowerBodyO2 == '<5' & XrayReport == 'Plethoric')))
```

```{r}
cpquery(bn.mle, event = (Disease == 'TGA'), evidence = ((LowerBodyO2 == '<5' & XrayReport == 'Plethoric')))
```

Based on the queries of diseases and the given evidence it is evident that the TGA causes lowerbodyo2 and xrayreport to be plethoric.


### (G)

given the lowerbodyo2<5 nd x-ray report to be oligaemic.

```{r}
cpquery(bn.mle, event = (Disease == 'Fallot'), evidence = ((LowerBodyO2 == '<5') & (XrayReport == 'Oligaemic') & (Grunting == 'no'))) 
```

```{r}
cpquery(bn.mle, event = (Disease == 'Lung'), evidence = ((LowerBodyO2 == '<5') & (XrayReport == 'Oligaemic') & (Grunting == 'no'))) 
```

```{r}
cpquery(bn.mle, event = (Disease == 'PAIVS'), evidence = ((LowerBodyO2 == '<5') & (XrayReport == 'Oligaemic') & (Grunting == 'no'))) 
```


```{r}
cpquery(bn.mle, event = (Disease == 'PFC'), evidence = ((LowerBodyO2 == '<5') & (XrayReport == 'Oligaemic') & (Grunting == 'no')))
```

```{r}
cpquery(bn.mle, event = (Disease == 'TAPVD'), evidence = ((LowerBodyO2 == '<5') & (XrayReport == 'Oligaemic') & (Grunting == 'no'))) 
```

```{r}
cpquery(bn.mle, event = (Disease == 'TGA'), evidence = ((LowerBodyO2 == '<5') & (XrayReport == 'Oligaemic') & (Grunting == 'no')))
```

There are two close likely diseases with Fallot being 45% and paivs being 35% 

### (h)


Baby julie is grunting with mild cardiac mixing.The disease is 

```{r}
cpquery(bn.mle, event = (Disease == 'Fallot'), evidence = ((Grunting == 'yes') & (CardiacMixing == 'Mild')))
```

```{r}
cpquery(bn.mle, event = (Disease == 'Lung'), evidence = ((Grunting == 'yes') & (CardiacMixing == 'Mild')))
```

```{r}
cpquery(bn.mle, event = (Disease == 'PAIVS'), evidence = ((Grunting == 'yes') & (CardiacMixing == 'Mild')))
```

```{r}
cpquery(bn.mle, event = (Disease == 'PFC'), evidence = ((Grunting == 'yes') & (CardiacMixing == 'Mild')))
```

```{r}
cpquery(bn.mle, event = (Disease == 'TAPVD'), evidence = ((Grunting == 'yes') & (CardiacMixing == 'Mild')))
```

```{r}
cpquery(bn.mle, event = (Disease == 'TGA'), evidence = ((Grunting == 'yes') & (CardiacMixing == 'Mild')))
```


Baby julie has the lung disease with 45% chance and others are shared between TGA,PFC and fallot.


Baby george is not grunting and with cardiacMixing.

```{r}
cpquery(bn.mle, event = (Disease == 'Fallot'), evidence = ((Grunting == 'no') & (CardiacMixing == 'Mild')))
```

```{r}
cpquery(bn.mle, event = (Disease == 'Lung'), evidence = ((Grunting == 'no') & (CardiacMixing == 'Mild')))
```

```{r}
cpquery(bn.mle, event = (Disease == 'PAIVS'), evidence = ((Grunting == 'no') & (CardiacMixing == 'Mild')))
```

```{r}
cpquery(bn.mle, event = (Disease == 'PFC'), evidence = ((Grunting == 'no') & (CardiacMixing == 'Mild')))
```

```{r}
cpquery(bn.mle, event = (Disease == 'TAPVD'), evidence = ((Grunting == 'no') & (CardiacMixing == 'Mild')))
```

```{r}
cpquery(bn.mle, event = (Disease == 'TGA'), evidence = ((Grunting == 'no') & (CardiacMixing == 'Mild')))
```

Baby julie has mostly at risk since baby george has 40% chance of contracting a most likely disease whereas julie has 45% chance of contracting a most likely disease.


